Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-10-28 paper/main.tex
paper

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations including  Plackett-Luce and Mallows
algorithms~\citep{???},\MANUEL{what can we cite from that body of research?: Ekhine will fix this.}
with the most prominent examples being estimation of distribution algorithms
(EDAs)~\cite{Ceberio2013}.  ACO may also be considered a type of EDA, since
it builds a probability distribution model from which solutions are sampled
with a bias towards the best solutions evaluated so far. The proposed UMM
drastically differs from EDAs since EDAs assume that function evaluations are
cheap and they can do many of them. On the contrary, UMM aims at learning as
much as possible with a limited number of function evaluations. Moreover,
unlike EDAs, UMM does not rely on a fixed-size population of solutions that
evolves over ``generations''. Instead, taking inspiration from Bayesian
optimizers, UMM considers a sample of permutations during the whole execution
that is increased by one new permutation at each iteration. Finally,
the balance between intensification and exploration is dynamically adapted by
controlling the variance of the probabilistic Mallows model.  \manuel{}{
  \begin{itemize}
  \item Who else has used Mallows model for black-box optimization? Ekhine: Va a mirarlo.
  \item Who else has used unbalanced borda for optimization? Ekhine va a añadir 2 frases.
  \end{itemize}
}
This paper structured as follows \MANUEL{Mío}

@@

\subsection{Unbalanced Borda}\label{sec:uborda}
We introduce now the core for our probabilistic algorithm, the uBorda algorithm. Following the convention of the community, the ranking returned by the uBorda algorithm will be denoted uBorda ranking. The uBorda algorithm is a recent generalization of Borda~\cite{???}.\MANUEL{X This paper is missing \ekhine{}{to be pocessed by arxiv}} In uBorda, we consider that each permutation in the sample $\sigma\in S$ has an associated weight $w(\sigma)$ where $w\colon S_n \to \mathbb{R^{+}}$. Then, the score of item $i \in [n]$ is defined as follows:
%
\begin{equation}\label{eq:uborda_score}
B(i) =  \sum_{t\in S}  w(\sigma_t) \sigma_t(i) \enspace,
\end{equation}
%
and, the same as Borda, orders the items $i \in [n]$ increasingly by their score.

Borda and uBorda are equivalent when $w(\sigma)$ is constant for every $\sigma$. However, in other case uBorda ranking will be closer to the rankings in $S$ with smaller value of $w(\sigma)$. Roughly speaking, it is equivalent to replicating in the sample the $\sigma$ with low values of $w(\sigma$).
Clearly, the choice of  $w(\sigma)$ depends on the relevant property of a ranking in a particular domain. For example, uBorda has been shown to have nice theoretical properties for the online estimation of the parameters of a sample with concept drift~\cite{???}.\MANUEL{X where? Do we miss a citation here?\ekhine{}{to be processed, arxiv} Maybe you can create a bibentry with the details that you know already.} 
%
% In this paper, we would like to replicate each permutation $\sigma$ a number of times that is inversely proportional to $f(\sigma)$. Therefore, we will use $w(\sigma)=\rho^{f(\sigma)}$ for a $\rho\in(0,1)$.
Interestingly, uBorda is computed in polynomial time.

@@

The \textbf{LOP} \ldots\EKHINE{Mío}

@@

For the LOP, we generated synthetic instances of size $n=20$, $m=200$, $\phi\in\{0.5,0.7,0.9\}$\MANUEL{what is $m$ here? We already use $\FEmax$ for the budget, but we could change it.}
Describir como se generan las intancias de LOP: Ekhine

Describir las LOPLIB instancias: Ekhine

@@

Over all the instances  we
see that the configuration $r_1 =0.1$, $r_2=0.9$ has good
performance\MANUEL{how do you define this? The best mean?}. From this
preliminary experiment, it is not obvious how the best settings are related to
problem features, such as permutation length ($n$). On the other hand, given
the sharp transitions in the plots, it is clear that further fine-tuning may
improve performance and uncover further patterns.  Although automatic parameter
tuning may likely provide even more fine-tuned settings, our goal here is to
understand how UMM performs in comparison to an existing Bayesian method (CEGO)
and automatically tuning the parameters of CEGO is not computationally
feasible.  For consistency, in the following sections, we run UMM with this
parameter configuration. Moreover, we recommend this configuration as the
current default of UMM.

@@

To better understand the behaviour of the algorithms, we record the fitness of
the solution evaluated at each step of each run and we average those values
over 10 runs with different random seeds on the same problem instance.  As an example, each line in Fig.~\ref{fig:lop_synth} 
shows the mean fitness of the solutions evaluated by either UMM or CEGO on one synthetic LOP
instance. The shaded area shows one standard deviation around the
mean. Ideally, each new evaluation will monotonically improve in fitness, since each solution evaluated helps to refine the model and leads to a
better solution being evaluated in the next step.  However, it is often the case that 
new solutions
are often worse than their predecessors. Nevertheless, in this example, there
is a clear overall improvement as more evaluations are performed. The minimum
values achieved in the plot up to a particular number of function evaluations
gives an estimation of the mean fitness of the best-so-far solution up to that
point in the run. The algorithms do not return the final solution shown but
rather the best one (minimum fitness) of each run.

@@

When we compare the behaviour of CEGO and
UMM on synthetic LOP instances (Fig.~\ref{fig:lop_synth}), we observe a clear
pattern independently of the uniformity of the instances. In particular, we see
a linear convergence to the optimal value for UMM.  This is due to the linear
decrease in the parameter $\theta$ (line~\ref{line:theta} in
Algorithm~\ref{alg:umm}), which imposes a linear variance decrease on the
internal probability model of the algorithm over the $m$ functions
evaluations. Overall, UMM found  the best solutions in all the repetitions.\MANUEL{in all runs UMM found a better solution? }

@@

On the other hand, \EKHINE{cambio el orden}
CEGO starts with a random sampling
of X\ekhine{}{10?} solutions, with are quite poor as expected, followed by building and
updating the Kriging model based new evaluations, which leads to an impressive
improvement in less than 50 evaluations. However, progress is %more or less 
halted at this point and the model seems to have trouble generating better
solutions. Lower uniformity values (e.g., $\phi=.5$) show worsening evaluations
despite the increase in solutions used to build the model. In summary, the
Kriging model clearly helps CEGO to quickly identify good solutions in very few
evaluations. However, for some reason, CEGO is not able to further improve
those solutions given more data. This may due to the model not being able to
produce accurate predictions or the underlying optimizer not being able to find
improved solutions using the model.\MANUEL{perhaps we should measure the
  prediction error and see if it grows or decreases. That would answer this
  question.}

@@

In the following paragraphs we analyze the result over real rata. Fig.~\ref{fig:loplib} shows the results on the instances available in LOPLIB~\citep{}. The distribution of this instances differs from the symthetic LOP instances, and thus so does the landscape. However, UMM shows a very similar performance: there is a (close to) linear decrease on the function evaluations of new solutions along iterations. This suggests that UMM very good fits the fitness landscape. Regarding CEGO, we can take similar conclussions for the first 50-100 iterations. However, the processes had to be halted after hitting the wall time, so no conlusion can be taken for further iterations. Overall, the improvement from the first (random) to the last solutions is 2-3 times larger in UMM than CEGO.

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (up to $1000$ or less).  UMM is
specially well-suited for budget-limited scenarios that are still
time-sensitive, e.g., when the overhead incurred by the optimizer should be no
more than a few minutes added to each fitness evaluation. As shown by our
results, a Bayesian optimizer such as CEGO may sometimes require more than hour
\EKHINE{me parece mas de CEGO q UMM y parece q UMM solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos  y luego que addemás es más rápido}
per fitness evaluation. The time required to train the surrogate-model of CEGO
also sharply increases with permutation length, hence, UMM would be a
computationally feasible alternative for relatively large problem sizes.

>>>>

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations, such as Mallows~\cite{gMallows} that have already been 
been used in optimization problems successfully,
with the most prominent examples being estimation of distribution algorithms
(EDAs)~\cite{Ceberio2013}, \cite{Ceberio2015}. ACO may also be considered a type of EDA, since
it builds a probability distribution model from which solutions are sampled
with a bias towards the best solutions evaluated so far. This works fall into at least one of these assumptions: function evaluations are cheap or the function evaluation is a white-box.

The proposed UMM
drastically differs from EDAs since EDAs assume that function evaluations are
cheap and they can do many of them. On the contrary, UMM aims at learning as
much as possible with a limited number of function evaluations. Moreover,
unlike EDAs, UMM does not rely on a fixed-size population of solutions that
evolves over ``generations''. Instead, taking inspiration from Bayesian
optimizers, UMM considers a sample of permutations during the whole execution
that is increased by one new permutation at each iteration. Finally,
the balance between intensification and exploration is dynamically adapted by
controlling the variance of the probabilistic Mallows model.

The proposed UMM is based on uBorda~\cite{irurozki2020}, a generalization of the well-known rank aggregation algorithm Borda~\cite{Borda1781}. Rank aggregation algorithms summarize in a single permutation the information of a sample of permutations $S$, and they can be seen as functions to compute  the \textit{mean} of sample $S$. The uBorda algorithm has been proposed to compute a \textit{weighted mean} of sample $S$. UBorda has been show to return the correct estimator with high probability on streaming contexts; in this paper, we show its quality on black-box optimization problems. \ekhine{}{revisa esto por favor, Manuel}

\manuel{}{
  \begin{itemize}
  \item Who else has used Mallows model for black-box optimization? Ekhine: Va a mirarlo.
  \item Who else has used unbalanced borda for optimization? Ekhine va a añadir 2 frases.
  \end{itemize}
}
This paper structured as follows \MANUEL{Mío}

@@

\subsection{Unbalanced Borda}\label{sec:uborda}
We introduce now the core for our probabilistic algorithm, the uBorda algorithm. Following the convention of the community, the ranking returned by the uBorda algorithm will be denoted uBorda ranking. The uBorda algorithm is a recent generalization of Borda~\cite{irurozki2020}. In uBorda, we consider that each permutation in the sample $\sigma\in S$ has an associated weight $w(\sigma)$ where $w\colon S_n \to \mathbb{R^{+}}$. Then, the score of item $i \in [n]$ is defined as follows:
%
\begin{equation}\label{eq:uborda_score}
B(i) =  \sum_{t\in S}  w(\sigma_t) \sigma_t(i) \enspace,
\end{equation}
%
and, the same as Borda, orders the items $i \in [n]$ increasingly by their score.

Borda and uBorda are equivalent when $w(\sigma)$ is constant for every $\sigma$. However, in other case uBorda ranking will be closer to the rankings in $S$ with smaller value of $w(\sigma)$. Roughly speaking, it is equivalent to replicating in the sample the $\sigma$ with low values of $w(\sigma$).
Clearly, the choice of  $w(\sigma)$ depends on the relevant property of a ranking in a particular domain. For example, uBorda has been shown to have nice theoretical properties for the online estimation of the parameters of a sample with concept drift~\cite{irurozki2020}.
%
% In this paper, we would like to replicate each permutation $\sigma$ a number of times that is inversely proportional to $f(\sigma)$. Therefore, we will use $w(\sigma)=\rho^{f(\sigma)}$ for a $\rho\in(0,1)$.
Interestingly, uBorda is computed in polynomial time.

@@

The \textbf{LOP} is defined as 
\begin{equation}\label{eq:lop}
  \min_{\sigma \in S_n} f(\sigma) = \sum_{i=1}^{n-1} \sum_{j=i+1}^n  a_{\sigma(i), \sigma(j)}
\end{equation}
where $[a_{i,j}]$ is a matrix of size $n \times n$.

@@

\paragraph{Datasets}

For the LOP, we generated synthetic instances as follows: sample 200 permutations from a MM of size $n=20$ and $\theta=\{-\log(.5), -\log(.7), -\log(.9)\}$, i.e., from peaked distributions to almost uniform distributions. Then, generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}(\sigma(i)<\sigma(j))$.

Moreover, we consider real LOP datasets from the LOPLIB repository. \footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}}
In particular, we consider the following types: IO, Random instances of type A and Random instances of type B.

@@

Over all the instances  we
see that the configuration $r_1 =0.1$, $r_2=0.9$ has good
performance: \ekhine{}{the rank of this parameter configuration among all the possible configurations is less than 4 on average for all the instances}. From this
preliminary experiment, it is not obvious how the best settings are related to
problem features, such as permutation length ($n$). On the other hand, given
the sharp transitions in the plots, it is clear that further fine-tuning may
improve performance and uncover further patterns.  Although automatic parameter
tuning may likely provide even more fine-tuned settings, our goal here is to
understand how UMM performs in comparison to an existing Bayesian method (CEGO)
and automatically tuning the parameters of CEGO is not computationally
feasible.  For consistency, in the following sections, we run UMM with this
parameter configuration. Moreover, we recommend this configuration as the
current default of UMM.

@@

To better understand the behavior of the algorithms, we record the fitness of
the solution evaluated at each step of each run and we average those values
over 10 runs with different random seeds on the same problem instance.  As an example, each line in Fig.~\ref{fig:lop_synth} 
shows the mean fitness of the solutions evaluated by either UMM or CEGO on one synthetic LOP
instance. The shaded area shows one standard deviation around the
mean. Ideally, each new evaluation will monotonically improve in fitness, since each solution evaluated helps to refine the model and leads to a
better solution being evaluated in the next step.  However, it is often the case that 
new solutions
are often worse than their predecessors. Nevertheless, in this example, there
is a clear overall improvement as more evaluations are performed. The minimum
values achieved in the plot up to a particular number of function evaluations
gives an estimation of the mean fitness of the best-so-far solution up to that
point in the run. The algorithms do not return the final solution shown but
rather the best one (minimum fitness) of each run.

@@

When we compare the behavior of CEGO and
UMM on synthetic LOP instances (Fig.~\ref{fig:lop_synth}), we observe a clear
pattern independently of the uniformity of the instances. In particular, we see
a linear convergence to the optimal value for UMM.  This is due to the linear
decrease in the parameter $\theta$ (line~\ref{line:theta} in
Algorithm~\ref{alg:umm}), which imposes a linear variance decrease on the
internal probability model of the algorithm over the $m$ functions
evaluations. Overall, \ekhine{UMM found  the best solutions in all the repetitions. }{the mean best solutions of UMM are better than the mean of CEGO regardless the uniformity of the instances. }

@@

On the other hand, % \EKHINE{cambio el orden}
CEGO starts with a random sampling
of 10 solutions, with are quite poor as expected, followed by building and
updating the Kriging model based new evaluations, which leads to an impressive
improvement in less than 50 evaluations. However, progress is %more or less 
halted at this point and the model seems to have trouble generating better
solutions. Lower uniformity values (e.g., $\phi=.5$) show worsening evaluations
despite the increase in solutions used to build the model. In summary, the
Kriging model clearly helps CEGO to quickly identify good solutions in very few
evaluations. However, for some reason, CEGO is not able to further improve
those solutions given more data. This may due to the model not being able to
produce accurate predictions or the underlying optimizer not being able to find
improved solutions using the model.\MANUEL{perhaps we should measure the
  prediction error and see if it grows or decreases. That would answer this
  question.}

@@

In the following paragraphs we analyze the result over real rata. Fig.~\ref{fig:loplib} shows the results on the instances available in LOPLIB.\footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}} The distribution of this instances differs from the symthetic LOP instances, and thus so does the landscape. However, UMM shows a very similar performance: there is a (close to) linear decrease on the function evaluations of new solutions along iterations. This suggests that UMM very good fits the fitness landscape. Regarding CEGO, we can take similar conclussions for the first 50-100 iterations. However, the processes had to be halted after hitting the wall time, so no conlusion can be taken for further iterations. Overall, the improvement from the first (random) to the last solutions is 2-3 times larger in UMM than CEGO.

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (up to $1000$ or less).  UMM 
\ekhine{}{has shown to perform statistically better or equal than CEGO, the state of the art algorithm, in the same instances as in the original paper of CEGO and some new instances. UMM}
is specially well-suited for budget-limited scenarios that are still
time-sensitive, e.g., when the overhead incurred by the optimizer should be no
more than a few minutes added to each fitness evaluation. As shown by our
results, a Bayesian optimizer such as CEGO may sometimes require more than hour
\EKHINE{me parece mas de CEGO q UMM y parece q UMM solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos  y luego que addemás es más rápido}
per fitness evaluation. The time required to train the surrogate-model of CEGO
also sharply increases with permutation length, hence, UMM would be a
computationally feasible alternative for relatively large problem sizes.
